---
title: "National Transit Database - 2015 Service Analysis"
author: "Ruzbeh Shokranian"
date: "February 2nd, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE)

library("ggplot2")
library("knitr")
library("dplyr")

##Load data in CSV format
ntd_service_data <- read.csv("Service.csv", sep=",", header=TRUE)

```

## National Transit Database Data Exploration

This document describes the analysis that was done on the National Transit 
Database (NTD) 2015 service data. The NTD program collects data from several
transit agencies in the United States and is the primary source for transit 
data.The specific dataset that will be analysed in this project correspond to 
2015 service data organized by agency, mode and type of service. The original 
dataset can be found on this link: 
https://www.transit.dot.gov/ntd/data-product/2015-service The excel file was 
downloaded and sheet "Annual Service Data by Mode" was saved in csv format and 
used in this project. The original excel dataset is attached with this project 
and it includes a sheet labeled "Data Dictionary" which describes each variable 
in the data.

The dataset includes information on several variables for many transit agencies.
Some of the variables include the number of annual trips, transit performance 
measures, revenue statistics and more. Starting the data exploration, a few 
plots will be made.

## Unlinked Trips by Mode

To start, annual unlinked trips by mode will be plotted. Unlinked trips are 
defined as the number of passengers that board or alight a transit vehicle, 
even if the boarding/alighting correspond to a transfer.

```{r mode_plot}
## Setting Mode Name
ntd_service_data$Mode_Name[ntd_service_data$Mode == "MB"] <- "Bus"
ntd_service_data$Mode_Name[ntd_service_data$Mode == "RB"] <- "BRT"
ntd_service_data$Mode_Name[ntd_service_data$Mode == "CC"] <- "Cable Car"
ntd_service_data$Mode_Name[ntd_service_data$Mode == "CB"] <- "Commuter Bus"
ntd_service_data$Mode_Name[ntd_service_data$Mode == "CR"] <- "Commuter Rail"
ntd_service_data$Mode_Name[ntd_service_data$Mode == "DR"] <- "Demand Response"
ntd_service_data$Mode_Name[ntd_service_data$Mode == "DT"] <- 
  "Demand Response Taxi"
ntd_service_data$Mode_Name[ntd_service_data$Mode == "FB"] <- "Ferryboat"
ntd_service_data$Mode_Name[ntd_service_data$Mode == "HR"] <- "Heavy Rail"
ntd_service_data$Mode_Name[ntd_service_data$Mode == "YR"] <- "Hybrid Rail"
ntd_service_data$Mode_Name[ntd_service_data$Mode == "IP"] <- "Inclined Plane"
ntd_service_data$Mode_Name[ntd_service_data$Mode == "LR"] <- "Light Rail"
ntd_service_data$Mode_Name[ntd_service_data$Mode == "MG"] <- "Monorail"
ntd_service_data$Mode_Name[ntd_service_data$Mode == "SR"] <- "Streetcar"
ntd_service_data$Mode_Name[ntd_service_data$Mode == "TB"] <- "Trolleybus"
ntd_service_data$Mode_Name[ntd_service_data$Mode == "VP"] <- "Vanpool"
ntd_service_data$Mode_Name[ntd_service_data$Mode == "JT"] <- "Jitney"
ntd_service_data$Mode_Name[ntd_service_data$Mode == "TR"] <- "Aerial Tramway"
ntd_service_data$Mode_Name[ntd_service_data$Mode == "PB"] <- "Publico"
ntd_service_data$Mode_Name[ntd_service_data$Mode == "AR"] <- "Alaska Railroad"
ntd_service_data$Mode_Name[ntd_service_data$Mode == "OR"] <- "OR"

## Making unlinked trips numeric
ntd_service_data$Unlinked.Passenger.Trips <- 
  as.numeric(gsub("\\,","",
                  as.character(ntd_service_data$Unlinked.Passenger.Trips)))

## Top unlinked trips by mode
p2 <- ggplot(ntd_service_data, aes(Mode))
p2 + geom_bar(aes(weight = Unlinked.Passenger.Trips))
```

Note that Bus (MB) and Heavy Rail (HR) correspond to the modes with the highest 
ridership. 

## State Analysis

The graph below shows annual unlinked trips grouped by state. The state with the
highest transit ridership corresponds to New York.

```{r state_plot}
ntd_unlinked_trips_by_state <- ntd_service_data %>% 
group_by(State) %>%
summarise(trips = sum(Unlinked.Passenger.Trips)) %>%
ungroup() %>%
arrange(desc(trips))

qplot(factor(State), data=ntd_unlinked_trips_by_state[1:10,], geom="bar", 
      fill=factor(State), weight = trips) +
  theme(legend.position = "none")

```

## City Analysis

The data includes the city in which each agency is located. First, some 
descriptive statistics of the number of unlinked trips per city and total 
passenger miles per 
city (defined as the number of miles that passengers travel) will be shown.

```{r city}

## Making passenger miles numeric
ntd_service_data$Passenger.Miles[is.na(ntd_service_data$Passenger.Miles)] <- 0
ntd_service_data$Passenger.Miles <- 
  as.numeric(gsub("\\,","",as.character(ntd_service_data$Passenger.Miles)))
## Making population numeric
ntd_service_data$Primary.UZA.Population <- 
  as.numeric(gsub("\\,","",
                  as.character(ntd_service_data$Primary.UZA.Population)))

ntd_data_by_city <- ntd_service_data %>% 
  group_by(City) %>%
  summarise(trips = sum(Unlinked.Passenger.Trips), passenger_miles = 
              sum(Passenger.Miles), city_population = 
              max(Primary.UZA.Population))  %>%
  ungroup() %>%
  arrange(desc(trips))

```
```{r city_statistics, echo = TRUE}
summary(ntd_data_by_city$trips)
summary(ntd_data_by_city$passenger_miles)

```

The range of both variables is very high, with the mean and the median having 
very different values. This is because there are several outliers in the data.
To better understand these variables the histograms below explore the trips and
passenger miles variables, respectively. A log transformation was used to 
improve visualization.

```{r city_histograms}
qplot(log(ntd_data_by_city$trips), 
      geom="histogram", bins = 10)

qplot(log(ntd_data_by_city$passenger_miles), 
      geom="histogram", bins = 10)

```

The graphs below 
presents the number of unlinked trips per city and total 
passenger miles per 
city, respectively. 
The graphs are shown for the top 10 cities.

```{r city_plot}

qplot(factor(City), data=ntd_data_by_city[1:10,], geom="bar", 
      fill=factor(City), weight = trips) +
  theme(legend.position = "none")
qplot(factor(City), data=ntd_data_by_city[1:10,], geom="bar", 
      fill=factor(City), weight = passenger_miles) +
  theme(legend.position = "none")

```

Both graphs above are very similar, indicating that there is a strong 
relationship 
between total unlinked trips and passenger miles. The scatter plot below shows 
that this relationship does in fact exist.

```{r city_scatter_plot}
## Scatter Plot: Total Trips X Total Passenger Miles
qplot(trips, passenger_miles, data=ntd_data_by_city, 
      main="Total Unlinked Trips X Total Passenger Miles", 
      xlab="Total Unlinked Trips", 
      ylab="Total Passenger Miles")
```

In the graph above, the point in the top right corner corresponds to New York 
City. Below, the same graph is plotted without NYC so it is possible to take a 
close look at the other cities. The graph indicates that although there is a 
relationship between both variables, there are a few cities where that 
relationship is weaker.

```{r city_scatter_plot_2}
## Scatter Plot: Total Trips X Total Passenger Miles
qplot(trips, passenger_miles, data=ntd_data_by_city[2:nrow(ntd_data_by_city),], 
      main="Total Unlinked Trips X Total Passenger Miles", 
      xlab="Total Unlinked Trips", 
      ylab="Total Passenger Miles")
```

It would be interesting to see how the population of the metropolitan area is 
related to transit ridership. The bar plot below presents the top 10 cities 
by population.

```{r city_population_plot}

qplot(factor(City), data=ntd_data_by_city[1:10,], geom="bar", fill=factor(City), 
      weight = city_population) +
  theme(legend.position = "none")

```

And here are some descriptive statistics of cities population:

```{r city_population_statistics, echo = TRUE}

summary(ntd_data_by_city$city_population)

```

And the city population histogram below:

```{r city_pop_histograms}
qplot(ntd_data_by_city$city_population, 
      geom="histogram", bins = 10)
```

And the scatter plot below shows how they are related:

```{r city_population_scatter_plot}
## Scatter Plot: Total Trips X City Population
qplot(city_population, trips, data=ntd_data_by_city, 
      main="Total Unlinked Trips X City Population", xlab="City Population", 
      ylab="Total Unlinked Trips")
```

## Vehicle Hours X Vehicle Miles

Vehicle Hours correspond to the number of hours that a vehicle travels while in 
service. Similarly, Vehicle Miles correspond to the number of miles that the 
vehicle travels while in service, The scatter plot below is an attempt to see if
there is a relationship between both variables. The plot applies a log 
transformation to each of these variables in an attempt to better represent the
results and to deal with outliers.

```{r vehicle_hours_miles}
ntd_service_data$Vehicle.Miles <- 
  as.numeric(gsub("\\,","",as.character(ntd_service_data$Vehicle.Miles)))

ntd_service_data$Vehicle.Hours <- 
  as.numeric(gsub("\\,","",as.character(ntd_service_data$Vehicle.Hours)))

ntd_service_data$Mode.VOMS <- 
  as.numeric(gsub("\\,","",as.character(ntd_service_data$Mode.VOMS)))

## Scatter Plot: Vehicle Hours X Vehicle Miles
p1 <- ggplot(ntd_service_data, aes(x = log(Vehicle.Miles), 
                                   y = log(Vehicle.Hours)))
p1 + geom_point(aes(color = Mode))
```

Each color in the graph corresponds to a different mode, and it appears that 
the relationship is different for each mode. Some of them seem to have a 
stronger relationship them others.

Continuing the exploration of vehicle related variables, the multivariate plot
below explores the relationshipt between three variables: Vehicle Hours, 
Vehicle Miles, and the Number of Vehicles, represented by diferent colors. As
expected all of them are related.

```{r vehicle_hours_miles_2}
## Scatter Plot: Vehicle Hours X Vehicle Miles X Number of Vehicles
p1 <- ggplot(ntd_service_data, aes(x = log(Vehicle.Miles), y = log(Vehicle.Hours)))
p1 + geom_point(aes(color = log(Mode.VOMS)))
```


To better present the relationship between Vehicle Hours & Miles for each mode,
a new variable will be created: the ratio of Vehicle Miles and Vehicle Hours.
The bar plot below presents this ratio for each mode.

```{r vehicle_miles_hours_ratio}
## Ratio Vehicle Miles / Vehicle Hours
ntd_service_data$Vehicle.Miles.Hours <- 
  as.numeric(ntd_service_data$Vehicle.Miles/
  ntd_service_data$Vehicle.Hours)

qplot(factor(Mode), data=ntd_service_data, geom="bar", fill=factor(Mode), 
      weight = Vehicle.Miles.Hours) +
  theme(legend.position = "none")
```

The bar plot above shows that the ratio is very different for each mode. Modes
such as Demand Response (DR) and Bus (MB) have a higher ratio. To confirm the 
correlation between both variables we will perform a Pearson's Test, which 
results in a 0.93 correlation.

```{r vehicle_miles_hours_pearson, echo=TRUE}

cor.test(ntd_service_data$Vehicle.Miles,ntd_service_data$Vehicle.Hours,
         method = "pearson")

```

## Commuter Rail

In this section Commuter Rail will be specifically analysed. The bar chart below
presents the top 10 commuter rail agencies in the US. Note that the 3 biggest 
serve the New York City metropolitan area.

```{r commuter_rail}
## Making speed numeric
ntd_service_data$Average.Speed..mi.hr. <- 
  as.numeric(as.character(ntd_service_data$Average.Speed..mi.hr.))

commuter_rail_subset <- filter(ntd_service_data, Mode == "CR")

## TOp 10 commuter rail agencies in the US
commuter_rail_by_agency <- commuter_rail_subset %>% 
  group_by(Name) %>%
  summarise(Trips = sum(Unlinked.Passenger.Trips)) %>%
  ungroup() %>%
  arrange(desc(Trips))

ggplot(data=commuter_rail_by_agency[1:10,], 
       aes(x=Name, y=Trips,fill=factor(Name))) +
  geom_bar(position="dodge", stat="identity") +
  coord_flip() +
  theme(legend.position = "none")

```

Another interesting visualization would be to compare each operator's ridership 
in respect to the mean. The graph below presents that:

```{r commuter_rail_mean_1}

## TOp 10 commuter rail agencies in the US in respect to the mean
ggplot(data=commuter_rail_by_agency[1:10,], 
       aes(x=Name, y=Trips/mean(commuter_rail_subset$Unlinked.Passenger.Trips)
           ,fill=factor(Name))) +
  geom_bar(position="dodge", stat="identity") +
  coord_flip() +
  theme(legend.position = "none")
```

The graph above indicates that NYC Commuter Rail operators have significantly 
higher ridership when compared to the mean ridership in the US. Below we present
the mean and the median for Commuter Rail Unlinked Trips:

```{r commuter_rail_mean_2, echo=TRUE}

mean(commuter_rail_subset$Unlinked.Passenger.Trips)
median(commuter_rail_subset$Unlinked.Passenger.Trips)

```

The mean and the mediam above are quite different, indicating that a few 
operators have very high ridership while most of them have low numbers. This can
be confirmed in the histogram below:

```{r commuter_rail_histogram}

qplot(commuter_rail_subset$Unlinked.Passenger.Trips, 
      geom="histogram", bins = 10)

```


The bar chart below presents the top 10 Commuter Rail Revenue Miles. That is the
number of miles travelled by trains while in revenue service.

```{r commuter_rail_revenue}
## TOp 10 COmmuter Revenue per mile 
revenue_commuter_rail_by_agency <- commuter_rail_subset %>% 
  group_by(Name) %>%
  summarise(Revenue_Mile = sum(as.numeric(Train..Revenue.Miles))) %>%
  ungroup() %>%
  arrange(desc(Revenue_Mile))

ggplot(data=revenue_commuter_rail_by_agency[1:10,], 
       aes(x=Name, y=Revenue_Mile
           ,fill=factor(Name))) +
  geom_bar(position="dodge", stat="identity") +
  coord_flip() +
  theme(legend.position = "none")
```

Comparing the graphs, New York City agencies are in top in all of them. However,
differences on the previous graph are lower, indicating that several commuter 
rail operators travel similar number of miles but carrying significantly 
different number of passengers.

## Bus

In this section the Bus mode will be analysed. The graph below presents the top 
10 bus operators in the US. 

```{r bus_operators}
## Making number of vehicles numeric
ntd_service_data$Mode.VOMS[is.na(ntd_service_data$Mode.VOMS)] <- 0

ntd_service_data$Mode.VOMS <- 
  as.numeric(gsub("\\,","",as.character(ntd_service_data$Mode.VOMS)))

## Making average passenger trip length numeric
ntd_service_data$Average.Passenger.Trip.Length..mi. <- 
  as.numeric(as.character(ntd_service_data$Average.Passenger.Trip.Length..mi.))
ntd_service_data$Average.Passenger.Trip.Length..mi.[
  is.na(ntd_service_data$Average.Passenger.Trip.Length..mi.)] <- 0 

bus_subset <- filter(ntd_service_data, Mode == "MB")

## Top 10 bus agencies in the US
bus_by_agency <- bus_subset %>% 
  group_by(Name) %>%
  summarise(trips = sum(Unlinked.Passenger.Trips), voms = sum(Mode.VOMS), 
            speed = mean(Average.Speed..mi.hr.), 
            trip_length = mean(Average.Passenger.Trip.Length..mi.)) %>%
  ungroup() %>%
  arrange(desc(trips))

ggplot(data=bus_by_agency[1:10,], 
       aes(x=Name, y=trips
           ,fill=factor(Name))) +
  geom_bar(position="dodge", stat="identity") +
  coord_flip() +
  theme(legend.position = "none")
```

The plots below explore the number of vehicles variable. The first graph 
presents the top 10 agencies grouped by number of vehicles while the second is a
histogram of agencies grouped by number of vehicles.

```{r bus_histogram}

## Plotting top 10 agencies by the number of vehicles
ggplot(data=bus_by_agency[1:10,], 
       aes(x=Name, y=voms
           ,fill=factor(Name))) +
  geom_bar(position="dodge", stat="identity") +
  coord_flip() +
  theme(legend.position = "none")

## Histogram presenting bus operators grouped by number of vehicles
qplot(bus_subset$Mode.VOMS, geom="histogram", bins = 40)

```

The trips and number of vehicles bar charts indicate that there is a 
relationship between both variables. Which can be confirmed in the scatter 
plot below:

```{r bus_scatter_plot}
## Scatter Plot: Total Trips X Total Passenger Miles
qplot(trips, voms, data=bus_by_agency, 
      main="Total Unlinked Trips X Total Vehicles", xlab="Total Unlinked Trips", 
      ylab="Total Vehicles")
```

And the correlation is confirmed with Pearson's test:

```{r bus_pearson, echo=TRUE}

cor.test(bus_by_agency$trips,bus_by_agency$voms,
         method = "pearson")

```

The bar plot below shows how average bus speed varies across agencies:

```{r bus_speed}

## Plotting top 10 agencies by the speed
ggplot(data=bus_by_agency[1:10,], 
       aes(x=Name, y=speed
           ,fill=factor(Name))) +
  geom_bar(position="dodge", stat="identity") +
  coord_flip() +
  theme(legend.position = "none")

```

And is there a relationship between the average bus speed and the average 
passenger trip length?

```{r bus_scatter_plot_2}
## Scatter Plot: Average Bus Speed X Average Trip Length
qplot(speed, trip_length, data=bus_by_agency, 
      main="Average Bus Speed X Average Trip Length", 
      xlab="Average Speed (miles/hour)", 
      ylab="Trip Length (Miles)")
```






## Schedule Vs. Reality

Another interesting variables to look at are the Scheduled Vehicle Revenue Miles
and Vehicle Revenue Miles. The first variable indicates the number of miles to 
be run by transit vehicles while in revenue service according to the schedule 
while the second variable presents the actual number of miles ran.

```{r schedule_reality_1}
## Scatter Plot: Scheduled.Vehicle.Revenue.Miles X Vehicle.Revenue.Miles
ntd_service_data$Scheduled.Vehicle.Revenue.Miles <- as.numeric(gsub("\\,","",
  as.character(ntd_service_data$Scheduled.Vehicle.Revenue.Miles)))

ntd_service_data$Vehicle.Revenue.Miles <- as.numeric(gsub("\\,","",
  as.character(ntd_service_data$Vehicle.Revenue.Miles)))

qplot(Scheduled.Vehicle.Revenue.Miles, Vehicle.Revenue.Miles, 
      data=ntd_service_data)
```

The scatter plot above shows that both variables are related, but as expected, 
the number of miles acutally ran is higher than scheduled for several transit 
agencies. 

## Final Plots and Summary

In this section, three graphs shown previously have been polished. 

The first plot shows the top 10 modes by unlinked trips. The plot shows that Bus
is the most used transit mode in the US, followed by Heavy Rail.

```{r final_mode_plot}
## unlinked trips by mode
ggplot(data=ntd_service_data, 
       aes(x=Mode_Name, y=Unlinked.Passenger.Trips
           ,fill=factor(Mode_Name))) +
  geom_bar(position="dodge", stat="identity") +
  coord_flip() +
  theme(legend.position = "none") +
  labs(x = "Mode", y = "Unlinked Trips", 
       title = "Annual Unlinked Trips by Mode") +
  scale_y_continuous(
    name="Unlinked Trips",
    labels = function(n) {format(n, big.mark=",", scientific = FALSE)})
```

The plot below shows the top 10 cities by unlinked trips. Note that New York 
City is has the most transit trips in the US with a significant difference when 
compared to the other cities.

```{r final_city_plot}

## Top 10 Cities by unlinked trips
qplot(factor(City), data=ntd_data_by_city[1:10,], geom="bar", fill=factor(City),
      weight = trips, main="Top 10 Cities by Unlinked Trips", xlab="City", 
      ylab="Unlinked Trips") + 
  theme(legend.position = "none") + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_fill_discrete(guide = guide_legend(title = "City")) +
  scale_y_continuous(
    name="Unlinked Trips",labels = 
      function(n) {format(n, big.mark=",", scientific = FALSE)})

```

The scatter plot below compares Schedules and Actual Revenue Miles. Scheduled 
Revenue Miles correspond to the amount of miles transit vehicles run as planned 
on the timetables, while Actual Revenue Miles correspond to the miles that the 
transit vehicles actually ran. A logarithm transformation was applied for both
variables in an attempt to better represent the data. The graph shows that the 
variables are strongly related 
and for most transit operators both variables have similar values.

```{r final_schedule_reality}
## Scatter Plot: Scheduled.Vehicle.Revenue.Miles X Vehicle.Revenue.Miles
qplot(Scheduled.Vehicle.Revenue.Miles, Vehicle.Revenue.Miles, 
      data=ntd_service_data, 
      main="Scheduled Vs. Actual Vehicle Miles", 
      xlab="Logarithm of Scheduled Vehicle Revenue Miles", 
      ylab="Logarithm of Actual Vehicle Revenue Miles") + 
  scale_x_log10() +
  scale_y_log10()
```

## Reflection

The NTD data provides several metrics that can be used to compare different 
transit operators across the US. This metrics can be use to evaluate the 
performance and quality of the service provided. The number of variables 
provided by the dataset is large and only a few of these variables were 
selected for analysis.

An interesting observation shown in this work is that the New York City 
Metropolitan area is the primary transit market in the US. Several of the 
graphs in this project have shown this. The city is first place in transit 
ridership with a signficant difference when compared to other cities.

One of the issues noticed while working with this data is that for some transit
operators the data is incomplete, for example, there are several agencies which
the passenger miles variable is zero, which implies that this data is not 
available.

The data used in this project correspond to a snapshot of 2015. The NTD also 
provides data for previous years, and for future work this data could be be used
to evaluate the performance of these transit agencies across time. Also, the 
data includes several variables that were unexplored, and other modes besides
Commuter Rail and Bus could be analysed.
